---
title: "HR Attrition Prediction and Retention Strategy Using Machine Learning"
author: "Mena Melaku"
date: "2025-03-28"
output: 
  html_document:
    theme: readable
    css: ../assets/styles.css
    toc: true
    toc_float: true
    number_sections: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(class)
library(caret)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)
library(dplyr)

```


# Introduction

This project builds a KNN model to predict whether an employee will leave or stay with their employer. 

The dataset includes features such as satisfaction, workload, salary, and department.

The model outputs:
- 0 = stayed
- 1 = left

# Step 1: Define the Problem

We are predicting whether an employee will leave.

Target variable:
- `left` (1 = left, 0 = stayed)

This is a binary classification problem.


# Step 2: Get and Clean the Data

Load the dataset and prepare it for modeling.

```{r load_data, echo=TRUE}
hr <- read.csv("../data/hrdata.csv")
```

remove unnecessary columns
```{r remove unnecessary index column, echo=TRUE}
hr$X <- NULL
```

```{r rename sales column to dept, echo=TRUE}
hr$dept <- hr$sales
hr$sales <- NULL
```

convert categorical variables to factors
```{r convert salary and dept to factors, echo=TRUE}
hr$salary <- as.factor(hr$salary)
hr$dept <- as.factor(hr$dept)
```

verify data structure
```{r view structure, echo=TRUE}
str(hr)
```
View summary statistics
```{r summary stats, echo=TRUE}
summary(hr)

colSums(is.na(hr))

summary(hr)
```

**Step 3:** Encode categorical variables
```{r categorical variables, echo=TRUE}
#model.matrix() creates dummy variables for factors
hr_dummies <- as.data.frame(model.matrix(~ . -1, data = hr))

# View the structure
str(hr_dummies)
```

Normalize all numeric values to range 0–1
```{r  normalize numeric values, echo=TRUE}
#Min-Max scaling function
minmax <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

#Apply to all columns
hr_norm <- as.data.frame(lapply(hr_dummies, minmax))

# View summary to confirm values are between 0 and 1
summary(hr_norm)
```


**Step 4:** Split the Data
```{r split the data, echo=TRUE}
set.seed(12345)  # for reproducibility (so you get the same split each time)
```

```{r identify test data, echo=TRUE}
#Pick 30% of rows randomly for testing
test_rows <- sample(1:nrow(hr_norm), 0.3 * nrow(hr_norm))

#Split into training and test sets
hr_test <- hr_norm[test_rows, ]
hr_train <- hr_norm[-test_rows, ]

#Save the target variable (left) separately for both sets
hr_test_y <- hr_test$left
hr_train_y <- hr_train$left

#Remove 'left' column from features (because we already saved it)
hr_test$left <- NULL
hr_train$left <- NULL
```

```{r check data, echo=TRUE}
#check data
nrow(hr_train)  # should be about 70% of the total 
nrow(hr_test)   # should be about 30% of the total 

head(hr_train_y)  # target variable 
head(hr_train)    # feature columns 
```

**Step 5:** Build the Model and Predict
```{r build knn, echo=TRUE}
#Run KNN 
knn_pred <- knn(train = hr_train,    # training data
                test = hr_test,      # testing data
                cl = hr_train_y,     # training labels (left column)
                k = 21)              # number of neighbors

#Check prediction output
summary(knn_pred)
```

# Step 6: Evaluate the Model
```{r evaluate knn, echo=TRUE}

# Generate the confusion matrix
confusionMatrix(as.factor(knn_pred), 
                as.factor(hr_test_y), 
                positive = "1")
```
                
## Business Impact Assumptions

To evaluate the financial impact of HR initiatives, we base our analysis on the following assumptions:

| Metric            | Value  | Notes |
|-------------------|--------|-------|
| Employees         | 14,999 |       |
| Attrition Rate    | 24%    | ~3,600 employees/year |
| Cost per Employee | $20,000| Recruiting, onboarding, lost productivity |
| Sensitivity       | 81%    | Correctly identifies employees who leave |
| Precision         | 83.8%  | % of predicted leavers that are correct |


We compare two potential HR initiatives:

| Initiative                 | Cost per Employee | Retention Success Rate |
|----------------------------|-------------------|-------------------------|
| **Low-Cost Recognition Program** | $500              | 15%                  |
| **High-Impact Career Development** | $1,000            | 25%                    |

These assumptions allow us to estimate:
- Number of leavers retained
- Total cost and ROI of interventions
- Savings from preventing attrition
- Waste/spend on false positives
- Overall effectiveness of each initiative

# Finanaical Impact 

Calculate the number of employees who leave, how many are identified by the model, and how many receive interventions.

For each initiative, compute:
- employees retained  
- total intervention cost  
- savings from retention  
- net savings  

```{r assumptions, echo=TRUE}
# Assumptions
total_employees <- 14999
attrition_rate <- 0.24
turnover_cost <- 20000
sensitivity <- 0.81
precision <- 0.838

# Model-predicted leavers
leavers_per_year <- total_employees * attrition_rate
identified_leavers <- leavers_per_year * sensitivity
adjusted_intervention_count <- identified_leavers / precision

# Initiative A: Recognition Program
recog_cost <- 500
recog_success <- 0.15
recog_retained <- identified_leavers * recog_success
recog_savings <- recog_retained * turnover_cost
recog_total_cost <- adjusted_intervention_count * recog_cost
recog_net_savings <- recog_savings - recog_total_cost

# Initiative B: Career Development
career_cost <- 1000
career_success <- 0.25
career_retained <- identified_leavers * career_success
career_savings <- career_retained * turnover_cost
career_total_cost <- adjusted_intervention_count * career_cost
career_net_savings <- career_savings - career_total_cost
```

Summarize the financial outcomes for each initiative, including retained employees, costs, savings, and net savings.

```{r results table, echo=TRUE}
results <- data.frame(
  Initiative = c("Recognition Program", "Career Development Program"),
  Retained_Employees = round(c(recog_retained, career_retained)),
  Adjusted_Intervention_Count = round(adjusted_intervention_count),
  Cost_per_Intervention = c(recog_cost, career_cost),
  Total_Intervention_Cost = round(c(recog_total_cost, career_total_cost), 2),
  Savings_from_Retention = round(c(recog_savings, career_savings), 2),
  Net_Savings = round(c(recog_net_savings, career_net_savings), 2)
)

results_clean <- results %>%
  mutate(
    Cost_per_Intervention = dollar(Cost_per_Intervention),
    Total_Intervention_Cost = dollar(Total_Intervention_Cost),
    Savings_from_Retention = dollar(Savings_from_Retention),
    Net_Savings = dollar(Net_Savings)
  )

kable(
  results_clean,
  caption = "Financial Impact of HR Retention Initiatives",
  align = "c"
) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE)
```

Visualize net savings for each initiative using a bar chart.
```{r display plot, echo=TRUE}

results_plot <- results %>%
  mutate(
    Initiative = factor(
      Initiative,
      levels = c("Career Development Program", "Recognition Program")
    )
  )

ggplot(results_plot, aes(x = Initiative, y = Net_Savings, fill = Initiative)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = dollar(Net_Savings, accuracy = 1)),
    vjust = -0.35,
    size = 5,
    fontface = "bold",
    color = "#00274C"
  ) +
  scale_y_continuous(
    labels = dollar_format(),
    expand = expansion(mult = c(0, 0.12))
  ) +
  scale_fill_manual(
    values = c(
      "Career Development Program" = "#FFCB05",
      "Recognition Program" = "#00274C"
    )
  ) +
  labs(
    title = "Net Savings by HR Initiative",
    subtitle = "Career Development delivers higher total savings, while Recognition is more cost-efficient",
    x = NULL,
    y = "Net Savings (USD)",
    caption = "Assumptions: 24% attrition rate, $20,000 turnover cost, 83.8% model precision"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 11, face = "bold", color = "#00274C"),
    axis.title.y = element_text(size = 12, face = "bold", color = "#00274C"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5, color = "#00274C"),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0, color = "gray40"),
    plot.margin = margin(15, 20, 15, 20)
  )
```

# Missed and Failed Interventions

Break down remaining turnover costs into two categories:

- missed leavers: employees who left but were not identified by the model  
- failed interventions: employees who were identified, received an intervention, but still left  
___

Use the attrition rate and model sensitivity to estimate total leavers, identified leavers, and missed leavers.

Then calculate how many identified employees still leave after each intervention.
```{r base values, echo=TRUE}
# Recalculate base values
total_employees <- 14999
attrition_rate <- 0.24
turnover_cost <- 20000
model_sensitivity <- 0.81

total_leavers <- total_employees * attrition_rate
identified_leavers <- total_leavers * model_sensitivity
missed_leavers <- total_leavers - identified_leavers  # false negatives

# Intervention failure (people still left after receiving help)
recog_success <- 0.15
career_success <- 0.25

recog_failed <- identified_leavers * (1 - recog_success)
career_failed <- identified_leavers * (1 - career_success)
```

Convert missed leavers and failed interventions into estimated turnover cost.
```{r convert to cost, echo=TRUE}
# Convert to cost
failures_df <- data.frame(
  Category = c("Missed Leavers", 
               "Recognition Program - Failed", 
               "Career Development Program - Failed"),
  Cost = round(c(
    missed_leavers * turnover_cost,
    recog_failed * turnover_cost,
    career_failed * turnover_cost
  ), 0)
)



total_leavers <- total_employees * attrition_rate
identified_leavers <- total_leavers * model_sensitivity
missed_leavers <- total_leavers - identified_leavers

recog_success <- 0.15
career_success <- 0.25

recog_failed <- identified_leavers * (1 - recog_success)
career_failed <- identified_leavers * (1 - career_success)

failures_df <- data.frame(
  Category = c("Missed Leavers", 
               "Recognition Program - Failed", 
               "Career Development Program - Failed"),
  Cost = round(c(
    missed_leavers * turnover_cost,
    recog_failed * turnover_cost,
    career_failed * turnover_cost
  ), 0)
)
```

Plot the estimated cost of missed leavers and failed interventions.
```{r graph, echo=TRUE}

# Create cleaner labels
failures_plot <- failures_df %>%
  mutate(
    Category = c(
      "Missed",
      "Recognition Failed",
      "Career Failed"
    )
  ) %>%
  mutate(
    Category = factor(
      Category,
      levels = c("Career Failed", "Recognition Failed", "Missed")
    )
  )

# Plot
ggplot(failures_plot, aes(x = Category, y = Cost, fill = Category)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = dollar(Cost, accuracy = 1)),
    vjust = -0.35,
    size = 5,
    fontface = "bold",
    color = "#00274C"
  ) +
  scale_y_continuous(
    labels = dollar_format(),
    expand = expansion(mult = c(0, 0.12))
  ) +
  scale_fill_manual(
    values = c(
      "Career Failed" = "#FFCB05",       # Maize
      "Recognition Failed" = "#00274C",  # Blue
      "Missed" = "#7F7F7F"               # Gray
    )
  ) +
  labs(
    title = "Cost of Missed and Failed Interventions",
    subtitle = "Most costs come from failed interventions rather than missed predictions",
    x = NULL,
    y = "Estimated Turnover Cost (USD)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold", color = "#00274C"),
    axis.title.y = element_text(size = 12, face = "bold", color = "#00274C"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5, color = "#00274C"),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.margin = margin(15, 20, 15, 20)
  )
```



Calculate return on investment for each initiative and compare results.

```{r roi, echo=TRUE}

# Calculate ROI
results$ROI <- round((results$Net_Savings / results$Total_Intervention_Cost) * 100, 1)
```


Calculate ROI for each initiative using net savings and total intervention cost.

Plot the results to compare return per dollar spent.
```{r roi plot, echo=TRUE}
# Clean labels
results_plot <- results %>%
  mutate(
    Initiative = c("Recognition", "Career Development")
  ) %>%
  mutate(
    Initiative = factor(
      Initiative,
      levels = c("Career Development", "Recognition")
    )
  )

# Plot
ggplot(results_plot, aes(x = Initiative, y = ROI, fill = Initiative)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(ROI, "%")),
    vjust = -0.35,
    size = 5,
    fontface = "bold",
    color = "#00274C"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  scale_fill_manual(
    values = c(
      "Career Development" = "#FFCB05",  # Maize
      "Recognition" = "#00274C"          # Blue
    )
  ) +
  labs(
    title = "ROI Comparison by HR Initiative",
    subtitle = "Recognition delivers higher return per dollar, despite lower total savings",
    x = NULL,
    y = "Return on Investment (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold", color = "#00274C"),
    axis.title.y = element_text(size = 12, face = "bold", color = "#00274C"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5, color = "#00274C"),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.margin = margin(15, 20, 15, 20)
  )
```
**Interpretation**  Failed interventions account for more cost than missed leavers. Compare these remaining costs before calculating ROI.

ROI Interpretation

Career Development has higher total savings, while Recognition has higher ROI.

Recognition is more efficient per dollar, while Career Development delivers greater total impact.

